Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2013-10-04 paper/paper.tex
more nudges

\begin{itemize}
 \item Nondeterministic values.
 \item Assumptions.
 \item Arbitrary width types.
\end{itemize}

So for example the programs in Fig.~\ref{fig:c-} are valid
C- programs, but the program in Fig.~\ref{fig:not-c-} is not, since it uses
a non-constant loop bound.

@@

Since each loop is bounded by a constant, and each recursive function call is
limited to a cosntant depth, a C- program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound $k$, then
a bounded model checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a C- program of
size $l$ and with largest loop bound $k$, a bounded model checker will
create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop free C- program
of size $O(s)$.  The safety problem for C- is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.

@@

To illustrate some of these points, we were able to implement a first
version of {\sc kalashnikov} in around 150 lines of code, which took
around two hours to write.  The resulting code was easy to modify,
optimise and debug.  The current version, which includes floating point
support, several optimisations and two model checking algorithms,
is still less than 1200 lines of code (600 lines of C- and 600 lines
of Python).

@@

The most successful optimizations we found were related to making fragments of
$\mathcal{L}$ and searching progressively larger fragments until a program
could be found.

\subsection{Limited Program Length}
The simplest restriction on programs is to only consider programs of a certain length.
We start by considering programs consisting of one instruction and progressively
increase this number.

\subsection{Word Width}
It is often the case that a program which satisfies the specification
will continue to satisfy it regardless of the word width of the machine it is
impemented on.  For example the program in Fig.~\ref{fig:bitvector-program}
isolates the least significant bit of an 8-bit word, but it will also
isolate the least significant bit of a 32-bit or indeed any sized word.

@@

The only wrinkle here is that sometimes a program we synthesize will contain
constants.  If we have synthesized a program for a machine with $k$-bit words,
the constants in the program will be $k$ bits wide.  To generalize the program
to a $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit
wide numbers from $k$-bit ones.  We have several strategies for this and
just try each in turn.  In the following, a subscript $n$ on a value means that
the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
word $a$, resulting in an $(lm)$-bit wide word.

@@

\subsection{Separate Constants}
$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
are implicit and a fresh register exists for each instruction to write its
output to.  A natural way to encode $\mathcal{L}$ instructions is to have an
opcode and two operands.  The opcode selects which instruction type is being
executed and the operands specify which registers should be operated on.
We might allow the operands to be either a register (i.e. a program argument
or the result of a previous instruction), or an immediate constant.

@@

\subsection{Explicit Search}
When we are working on an $\mathcal{L}$-machine with a small word size and
a compact program encoding, our state space is very small.
We can quickly check the validity of the {\sc synth} and {\sc verif} formulae
by simply enumerating all of the inputs and programs, then explicitly
executing each program on each input.

>>>>

\begin{itemize}
 \item Nondeterministic values.
 \item Assumptions.
 \item Arbitrary-width types.
\end{itemize}

For example, the programs in Fig.~\ref{fig:c-} are valid
C- programs, whereas the program in Fig.~\ref{fig:not-c-} is not, since it uses
a non-constant loop bound.

@@

Since each loop is bounded by a constant, and each recursive function call is
limited to a cosntant depth, a C- program necessarily terminates and in
fact does so in $O(1)$ time.  If we call the largest loop bound~$k$, then
a bounded model checker with an unrolling bound of $k$ will be a complete
decision procedure for the safety of the program.  For a C- program of
size $l$ and with largest loop bound~$k$, a bounded model checker will
create a SAT problem of size $O(lk) = O(l)$.  Conversely, a SAT problem
of size $s$ can be converted trivially into a loop-free C- program
of size $O(s)$.  The safety problem for C- is therefore NP-complete,
which means it can be decided fairly efficiently for many practical
instances.

@@

To illustrate some of these points, we were able to implement a first
version of {\sc kalashnikov} in around 150 lines of code, which took around
two hours to write.  The resulting code was easy to modify, optimise and
debug.  The current version, which includes support for floating-point
arithmetic, several optimisations and two model checking algorithms, is
still less than 1200 lines of code (600 lines of C- and 600 lines of
Python).

@@

The most successful optimizations we found were related to defining
fragments of $\mathcal{L}$ and searching progressively larger fragments
until a program is found.

\subsection{Limited Program Length}

The simplest restriction on programs is to only consider programs of a certain length.
We start by considering programs consisting of one instruction and progressively
increase this number.

\subsection{Word Width}

It is often the case that a program which satisfies the specification will
continue to satisfy it regardless of the word width of the machine it is
impemented on.  For example, the program in Fig.~\ref{fig:bitvector-program}
isolates the least-significant bit of an 8-bit word, but it will also
isolate the least-significant bit of a 32-bit word or indeed of a word of
any size.

@@

The only wrinkle here is that sometimes a program we synthesize will contain
constants.  If we have synthesized a program for a machine with $k$-bit words,
the constants in the program will be $k$ bits wide.  To generalize the program
to an $n$-bit machine (with $n > k$), we need some way of deriving $n$-bit-wide
numbers from $k$-bit ones.  We have several strategies for this and
just try each in turn.  In the following, a subscript $n$ on a value means that
the value is an $n$-bit word.  The dot operator $a_l \cdot b_m$ denotes
concatenation, resulting in an $(l+m)$-bit wide word.  A superscript on a value
means duplication, so $a_l^m$ means concatenating together $m$ copies of the $l$-bit
word $a$, resulting in an $(lm)$-bit wide word.

@@

\subsection{Separate Constants}

$\mathcal{L}$ is a SSA, three address instruction set.  Destination registers
are implicit and a fresh register exists for each instruction to write its
output to.  A natural way to encode $\mathcal{L}$ instructions is to have an
opcode and two operands.  The opcode selects which instruction type is being
executed and the operands specify which registers should be operated on.
We might allow the operands to be either a register (i.e. a program argument
or the result of a previous instruction), or an immediate constant.

@@

\subsection{Explicit Search}

When we are working on an $\mathcal{L}$-machine with a small word size and
a compact program encoding, our state space is very small.
We can quickly check the validity of the {\sc synth} and {\sc verif} formulae
by simply enumerating all of the inputs and programs, then explicitly
executing each program on each input.
